---
title: 'Patient Access to Ambulatory Care Services in Lilongwe Urban: Area 25 Health
  Center'
author: "Malongo, Robert"
date: "7/19/2018"
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

# Data

## Introduction
* This is an analysis of healthcare services usage at Area 25 Health Center in Lilongwe, Malawi. 
* We carry out a brief descriptive analysis using frequency tables and graphical illustrations in assessing primary, secondary, and additional diagnosis of patients at the Out-Patient Department (OPD).
* This data was stored in and pulled from  MySQL Database Management System.
* This analysis is based on data from 12/21/2011 to 10/08/2013.

```{r loadrelevantpackages,echo  =TRUE, message=FALSE, warning=FALSE}

# Load Relevant packages
library("dplyr")
library("stringr")
library("tidyr")
library("ggplot2")
library("stringr")
library("lubridate")
library("gridExtra")
library("ggmap")
library("leaflet")
library("readr")

# Set working Directory
setwd("~/Desktop/summer_2018_projects")

# Import the SQL file data
concept <- read.delim("./sql_query/conceptnames.txt", header = FALSE)
```

## Analysis Procedure

* Click the code button on the bottom right of this section to review the analysis before proceeding to the next section

```{r dataanalysis,echo = TRUE,  message=FALSE, warning=FALSE, include = TRUE}
# Clean data   --------------------------------------------------------------

# Rename variables
df1_concept  <-  concept %>% 
  rename(obs_id = V1,
         concept_id  = V2,
         person_id = V3,
         obs_datetime = V4,
         value_coded = V5,
         voided = V6,
         name = V7,
         gender = V8,
         birthdate = V9,
         birthdate_estimated = V10,
         address1 = V11,
         address2 = V12,
         city_village = V13,
         county_district =V14)

#  Modify data types
df1_concept <- df1_concept %>% 
  mutate(name = as.character(name),
         obs_datetime = ymd_hms(as.character(obs_datetime)),
         birthdate = ymd(as.character(birthdate)),
         address1 = as.character(address1),
         address2 = as.character(address2),
         city_village  = as.character(city_village),
         county_district = as.character(county_district),
         age = round((obs_datetime - as.POSIXct.Date(birthdate))/dyears(),0))

# Assign NAs
df1_concept$address1[df1_concept$address1  %in% c ("\\N", "")] <- NA
df1_concept$address2[df1_concept$address2  %in% c ("\\N", "")] <- NA
df1_concept$city_village[df1_concept$city_village  %in% c ("\\N", "")] <- NA
df1_concept$county_district[df1_concept$county_district  %in% c ("\\N", "")] <- NA

# Recode city village column
df1_concept$city_village[df1_concept$city_village %in% c ("Area25a","Area 25 a", "Area 25a", "25a" , "Area 25 A ",  
                                                         "Area 25a Houses", "Area 25a Location", " Area 25 A") ] <- "Area 25 A"
df1_concept$city_village[df1_concept$city_village %in% c ("Area25b", "Area 25 B ", "Area 25 b", "Area 25b", "25b", "Area25b near cindillela", "Area2b") ] <- "Area 25 B"
df1_concept$city_village[df1_concept$city_village %in% c ("Area25c","Area 25 c", "Area 25 C ",  "Area 25c", "Are25c", "Area5c", "Area2c","Area 2c", "Area 25c Housing", "Area 25c Location", "Area925c", "Area  25c", " Area 25 C", "25c") ] <- "Area 25 C"
df1_concept$city_village[df1_concept$city_village %in% c ("Area25","25", " 25", "25 ", "Area  25", "Area25 ", "Area 25 ") ] <- "Area 25"

df1_concept$city_village[df1_concept$city_village %in% c ("Collage", "Ttc","Lttc", "Ltcc", "Tt", "Lilongwe teachers college", "Ltc", "Lct", "College")] <- "LL TTC"

df1_concept$city_village[df1_concept$city_village %in% c ("Sec1","Sector 1", "25sector1")] <- "Sector 1"
df1_concept$city_village[df1_concept$city_village %in% c ("Sec2", "Sectar 2", "Secter 2" , "Sector2")] <- "Sector 2"
df1_concept$city_village[df1_concept$city_village %in% c ("Sec3",  "Secter 3" ,"Sector3", "Setor3" )] <- "Sector 3"
df1_concept$city_village[df1_concept$city_village %in% c ("Sec5","Secter 5", "Secter  5", "Sector5")] <- "Sector 5"
df1_concept$city_village[df1_concept$city_village %in% c ( "Sec 6",  "Sec6", "Secter 6" ,  "Sector6")] <- "Sector 6"
df1_concept$city_village[df1_concept$city_village %in% c ( "Area 25 A sec7", "Area 25 sec 7", "Area 25 sector 7", "Area25 sector 7",
                                                           "Area 25sec7", "Area 25sector 7", "Sector 7", "7", "Area 25  sec 7", "Secter7", "Se7","Sector7", "Secto 7","Secto 7" ,"Secter 7" ,  "Sec7", "Secor 7" ,"Sec 7","Sctor 7","Sectr7", "Setor 7","Sec78")] <- "Sector 7"
df1_concept$city_village[df1_concept$city_village %in% c ("Sector8", "Sectr8", "Secter8" , "Secter 8","Sec8")] <- "Sector 8"
df1_concept$city_village[df1_concept$city_village %in% c ("Sector9", "Sec9")] <- "Sector 9"


df1_concept$city_village[df1_concept$city_village %in% c ("25sungwi", "Nsunwi", "Nsungwi", "Sungwi Trading centre", "Sugwi", "Nsungwi")] <- "Sungwi"
df1_concept$city_village[df1_concept$city_village %in% c ("Mgomani", "Gomani", "Ngoman", "Ngomani ", "Ngomanii", "2ngomani", "Ngoomani", "Ngomn","Ngomaon",
                                                          "Ngomane", "Ngomam", "Ngoamani", "Ngmani", "Ng'oman","Ngommani","Ngomani---", "Ngomana", "Ngomain","Ngoani","Ngoaman","Ngman", "Ngomani village",
                                                          "Ngomamn","Mgoani", "Mgmani","Momani",  "Mgolman", "Mgolma", "Mgmani", "Ngomai","Ngoan", "Ngomni","Ngnman","Mgoman","Mgnomani","Ngomaoni", "Ngomami" , 
                                                          "Ngoma","Ngoami","Nogmani","Nomani", "Ngmna", "Ngamani","Amgona",  "Mgomni", "Mgomana", "Mgolani", "Mngomani",  "Mmgomani")] <- "Ngomani"

df1_concept$city_village[df1_concept$city_village %in% c ("Mgona village", "Mgomna","Mg'oma","Mgonna","Mgona c","Mgoma","Mgna",
                                                          "Mngona","Mgoa", "Mmgona" , "Mhona", "Mgonaa",  "Mgona b",  "Mgona a" , "Mgon",  "Ngona c",  "Ngona", "Ngon",  "Ng'oma","Mgoni","Mgona "  )] <- "Mgona"

df1_concept$city_village[df1_concept$city_village %in% c ("Mwmbakathu", "Mwmbakanthu","Mwambwakanthu","Mwambkathu","Mwambkanthu",
                                                          "Mwambathu","Mwambanthu", "Mwambalathu", "Mwambaknthv", "Mwambaknthu",
                                                          "Mwambakatu", "Mwambakathu", "Mwambakantu", "Mwambakanthv", "Mwambakanthuu",
                                                          "Mwambakanthu","Mwambakanhu",  "Mwambaka" ,"Mwamba","Mwambwakanthu","Mwamakathu", "Mwakathu", "Mwayakanthu","Mwayambakanthu","Mwabakanthu" 
                                                          )] <- "Mwambakathu"


df1_concept$city_village[df1_concept$city_village %in% c ("Kulyani", "Kuliyni", "Kuliyini",  "Kuliyanii", "Kuliyani  vge",
                                                          "Kuliyana",  "Kuliyan", "Kuliyam", "Kuliya", "Kulilani",  "Kuliani",
                                                          "Kulian",  "Kuleyan", "Kliyani", "Ktliani")] <- "Kuliyani"

df1_concept$city_village[df1_concept$city_village %in% c ("Khanyiji","Kanyinji", "Khanyinji","Khanyiji" ,"Khanyeji", "Khanxyinji", "Khanjinji",
                                                          "Khanjinje",  "Khaninyi", "Khaninji", "Khanenji","Khaminji","Khamenji",
                                                          "Khainji", "Kayinv", "Kayinji", "Kayimvi",  "Mkhanyinji", "Mkhaninji", "Mkaninji",  "Nkhaninji", 
                                                          "Nkhanyinji",  "Nkhaninj", "Nkaninji" , "Nkhanji"  )] <- "Khanyinji"

df1_concept$city_village[df1_concept$city_village %in% c ("Area 18 ", "Are18", "Aea18","18"  )] <- "Area 18"

df1_concept$city_village[df1_concept$city_village %in% c ( "Area 18 B Location" )] <- "Area 18 B"

df1_concept$city_village[df1_concept$city_village %in% c ( "23", " 23", "Area 23 Location"," Area 23", "Area 23 Trading Centre "
 )] <- "Area 23"

df1_concept$city_village[df1_concept$city_village %in% c ("49", " 49","49 ","Area49","Area 49", "Aea49", "Are49", "Area 49 Houses", "Area 49 Trading Centre", 
                                                          "Area 49 proper", "Areaa49", "49dobay", "Bagdad", "Bagidadi Houses", "Bakidadi", "Area49dubai", 
                                                          "Shire (A49)", "Dubai", "  49", " Area 49", "Area  49", "Gulliver (A49)") ] <- "Area 49"

df1_concept$city_village[df1_concept$city_village %in% c ("Area  50", "50", " 50","50 ","A50", "Area50", "Aa50", " Area 50", "Aea50", "Areaa50", "Areah50",
                                                          "Arrea50", "Area50b", "Area550" ,"500") ] <- "Area 50"

df1_concept$city_village[df1_concept$city_village %in% c ("51", " 51", "51 ", "Area51","Area 51")] <- "Area 51"

df1_concept$name[df1_concept$name %in% c ("Respiratory tract infections, recurrent (sinusitis, tonsilitus, otitis media, pharyngitis)")] <-"RTI, recurrent"

df1_concept$age[df1_concept$age < 0] <- NA

# Manipulate data ---------------------------------------------------------

# Remove voided obs
df2_concept <- df1_concept %>% 
  filter(voided == 0) %>% 
  
# Keep one obs for each concept_id; note loss of obs
  group_by(obs_id, value_coded) %>% 
  slice(1) %>% 
  ungroup()

# Create location .csv export file
coord_city_village<- df2_concept %>% 
  select(city_village) %>% 
  unique.data.frame() 
write.csv(coord_city_village, "./output/area_25/coord_city_village.csv")

coord_address1<- df2_concept %>% 
  select(address1) %>% 
  unique.data.frame()
write.csv(coord_address1, "./output/area_25/coord_address1.csv")
```

#  Overall Analysis

* Here we look at all observations pulled from MySQL in the specified time frame

## Cases by Location

* Here are the top twenty area of residence for patients at Area 25 Health Center
* Most patients come Area 25 locations (A, B, an C), Ngona slums and Ngomani
* A significant number of patients come from as far as Mchezi and Chimoka

```{r plot1, echo=TRUE, message=FALSE, warning=FALSE, fig.height= 5, fig.width=15}
# Prelim Analysis ---------------------------------------------------------

df_sum <- df2_concept %>% 
  select(city_village,name) %>% 
  group_by(city_village) %>% 
  summarise(`Number of Cases` =n()) %>% 
  ungroup() %>% 
  arrange(desc(`Number of Cases`)) %>% 
  head(20)

# Plo1 1: Number of Cases by Location
ggplot(data = df_sum, aes(x = reorder(city_village,`Number of Cases`),  y = `Number of Cases`, fill = `Number of Cases`)) + 
  geom_col() +
  scale_fill_gradient(low = "green", high = "red") + 
  ggtitle("OPD Cases: Area 25 Health Center") +
  labs(caption  ="*Based on data from 12/21/2011 to 10/08/2013") +
  xlab("Patient Residence") +
  ylab("Number of Cases") 
```

## Cases by Age
* It is not surprising that most patients are below the age of five
* It is quite interesting that there are also many patients who are in their early 20s
```{r plot2, echo=TRUE, message=FALSE, warning=FALSE, fig.height= 7, fig.width=15}
# Plot 2: Number of Cases by Age
hist(x = df2_concept$age, col = "beige", xlab = "Age", main = "OPD Cases Area 25 Health Center: Age Distribution")
```


## Diagnosis by Location

* Here are the top ten diagnosis at the health center distributed by location
* Acute Respiratory Infection, Malaria, and Musculoskeletal pain are the three most common diagnosis 
* The three are the most common diagnosis at Area 25 and Ngona

```{r plot3, echo=TRUE, message=FALSE, warning=FALSE, fig.height= 5, fig.width=15}
# Top 10 Diagnosis
x_name_freq <- data.frame(table(df2_concept$name)) 

df_sum2<- df2_concept %>% 
  select(city_village,name) %>% 
# Select sample locations
  filter(city_village %in% c("Area 25 A", "Area 25 B", "Area 25 C", "Ngomani", "Mgona", "Area 50", "Area 49", 
                             "Chimoka", "Mchezi", "Area 25", "Chatata"),
# 10 most common diagnosis
         name %in% c("Acute respiratory infection", "Malaria", "Musculoskeletal pain","Respiratory Tract Infections, recurrent",
                     "Other skin condition", "Sepsis", "Diarrhoea diseases", "Abscess", 
                     "Dental caries", "Gastroenteritis")) %>% 
  group_by(city_village, name) %>% 
  summarise(`Number of Diagnosis` =n()) %>% 
  ungroup() %>% 
  mutate(`Percent of Total Cases` = round((`Number of Diagnosis`/sum(`Number of Diagnosis`)*100),2)) %>% 
  arrange(desc(`Number of Diagnosis`)) %>% 
  ungroup()

# Plot 2: Diagnosis  & Location
ggplot(df_sum2, aes(x=city_village, y=name, fill = `Percent of Total Cases`)) + geom_tile() +
  scale_fill_gradient(low = "yellow", high = "red") +
  ggtitle("Common Diagnosis: Area 25 Health Center") +
  xlab("Patient Residence") +
  ylab("Name of Diagnosis") 
```

## Diagnosis by Gender

* There are more cases of Acute Respiratory Infection among females than males

```{r plot4, echo=TRUE, message=FALSE, warning=FALSE, fig.height= 8, fig.width=15}
# Plot 4: Gender  & Name
 df_sum3<- df2_concept %>% 
   select(name, gender) %>% 
# 10 most common diagnosis
   filter(name %in% c("Acute respiratory infection", "Malaria", "Musculoskeletal pain","RTI, recurrent",
                         "Other skin condition", "Sepsis", "Diarrhoea diseases", "Abscess", 
                         "Dental caries", "Gastroenteritis")) %>% 
   group_by(name, gender) %>% 
   mutate(n = n()) %>% 
   ungroup() %>% 
   unique.data.frame()
 
 # Plot 4: 10 Top Diagnosis & Gender
 ggplot(df_sum3, aes(x= name, y= n, fill = n)) + geom_col()+
   scale_fill_gradient(low = "green", high = "red") + facet_wrap(~gender, nrow = 2) +
   ggtitle("Common Diagnosis by Gender: Area 25 Health Center") +
   xlab("Name of Diagnosis") +
   ylab("Number of Cases") 
 
 # Remove Some data frames
rm(df1_concept, df_sum, df_sum2, df_sum3, x_name_freq, coord_address1, coord_city_village, concept)
```

# Malaria Analysis
* Here we focus on Malaria diagnosis only

## Cases by Location

* Very similar trend to the overall analysis
```{r plot5, echo=TRUE, message=FALSE, warning=FALSE, fig.height= 7, fig.width=15}
df_malaria <- df2_concept %>% 
  filter(name == "Malaria") %>% 
  mutate(age_group =ifelse(age<5, "Under 5s",
                           ifelse(age>=5 & age <= 14, "5-14", 
                                  ifelse(age >14 & age <= 49, "15-49",
                                         ifelse(age > 49 & age <= 69 , "50-69","70 +")))))
 
df_sum_malaria <- df_malaria %>% 
  select(city_village,name) %>% 
  group_by(city_village) %>% 
  summarise(`Number of Cases` =n()) %>% 
  ungroup() %>% 
  arrange(desc(`Number of Cases`)) %>% 
  head(20)

# Plot 5: Number of Cases by Location
ggplot(data = df_sum_malaria, aes(x = reorder(city_village,`Number of Cases`),  y = `Number of Cases`, fill = `Number of Cases`)) + 
  geom_col() +
  scale_fill_gradient(low = "green", high = "red") + 
  ggtitle("OPD Malaria Cases: Area 25 Health Center") +
  labs(caption  ="*Based on data from 12/21/2011 to 10/08/2013") +
  xlab("Patient Residence") +
  ylab("Number of Cases") 
```

## Cases by  Age

* Also shows very  similar trend to the overall analysis
* Disproportionately more cases among under five kids
* More female cases in all the five age groups

```{r plot6, echo=TRUE, message=FALSE, warning=FALSE, fig.height= 7, fig.width=15}
# Plot 6: Age Distribution
hist( x = df_malaria$age, col = "beige", xlab = "Age", main = "OPD Malaria Cases at Area 25 Health Center: Age Distribution")

# Age group
df_sum_malaria_2 <- df_malaria %>% 
  group_by(gender, age_group) %>% 
  summarise(n = n()) %>% 
  spread(key = gender,  value = n) %>% 
  na.omit() %>% 
  arrange(desc(M))


knitr::kable(df_sum_malaria_2 , align = "l")
```

# Spatial Analyisis
* This section provides an in depth Geo Spatial analysis of the relationship between the number of healthcare service providers and the population in their catchment area.

## Total Population: 2008
* Please slide left, right, or zoom in the leaflet map for better visualization
* You can also click on the markers to access more information about the location
* Area 25, Area 56 (Chinsapo and Likuni), and Area 36 are the three most populous locations in Lilongwe urban
 
```{r plot7, echo=TRUE, message=FALSE, warning=FALSE, fig.height= 8, fig.width=12}
# Import LL urban coordiantes file
ll_coord<- read_csv("./shape_file/ll_coordinates.csv") %>% 
  select(-X1)

# Customize Marker Colors
get_color <- function(ll_coord) {
  sapply(ll_coord$total_pop, function(total_pop) {
    if(total_pop <= 10000) {
      "green"
    } else if(total_pop <= 30000) {
      "orange"
    } else {
      "red"
    } })
}

# Customize icon colors
icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'black',
  library = 'ion',
  markerColor = get_color(ll_coord)
)

# Create bin formats
bins <- c(0, 10000, 30000, Inf)
pal <- colorBin(c("green", "orange", "red"), domain = ll_coord$total_pop, bins = bins)

# Plot Leaftet Map 
leaflet(data = ll_coord)%>% 
  setView(lat = -13.962612, lng =  33.828119 , zoom = 12) %>% 
  addTiles() %>% 
  addAwesomeMarkers(lat = ~lat, lng = ~lon, popup = ~ as.character(total_pop), label = ~AreaName, icon = icons) %>% 
  addLegend(
    pal = pal,
    values = ~total_pop,
    opacity = 1,
    title = "Location Population",
    position = "topleft")
```

## Health Service Providers
* Please slide left, right, down, or zoom in the map for better visualization
* Also, click on the markers for more information
* There are few healthcare providers in comparison to the population of the surrounding location
* Area 56, Area 57 (Ntandile and Mtsiriza), and Area 36 are some of the most populous locations without a public healthcare provider within close proximity

```{r plot8, echo=TRUE, message=FALSE, warning=FALSE, fig.height= 8, fig.width=12}
# Import LL health center coordinates
ll_health_centr <- read_csv("./shape_file/ll_health_centr.csv") %>% 
  select(-X1) %>% 
  mutate(name = str_sub(name, start = 1, -19))

leaflet(data = ll_health_centr) %>% 
  setView(lat = -13.97000, lng =  33.828119 , zoom = 12) %>%
  addTiles() %>% 
  addAwesomeMarkers(lat = ~lat, lng = ~lon, popup = ~hos_type, 
                    label = ~name, labelOptions = (labelOptions(noHide = T)))

```